# R

## Assignment

For assigning a value to a variable, i use `<-` unless the variable is a parameter or an argument - then i use `=`:

```{r label = "assignment"}
variable <- "value"

name_of_function <- function(parameter = "value") {
  
}

```

## Debugging

### Printing out a value

```{r label = "print"}
print(2)

```

### Datatype

For figuring out the datatype [@zach_2021_how]:

```{r label = "datatype"}
class(2)

```

### Datatype and value

```{r label = "str"}
str(2)

```

### Is the value not set?

```{r label = "is-null"}
is.null(variable)

```

### Stopping execution

```{r label = "stop", eval = FALSE}
stop("error message")

```

## Numbers

### Converting a string to a number

```{r label = "as-numeric"}
number = "3"
str(number)

numeric_number <- as.numeric(number)
str(numeric_number)

```

### Sequencing

```{r label = "seq"}
seq(0, 14, 2)

```

### Rounding

#### Normal rounding

```{r label = "round"}
round(123.456, digits = 1)

```

#### Significant digits

For printing only significant digits, there's the function `signif` which takes two parameters: the initial number `x` and the number of significant digits `digits`:

```{r label = "signif"}
number <- 123.456

number_with_significant_digits <- signif(x = number, digits = 4)
print(number_with_significant_digits)

```

## Time

### Converting to time

```{r label = "as-date"}
string_of_time = "2020.7.08 16:43:59"
str(string_of_time)

librarian::shelf(c(
  "lubridate"
))

string_of_time_as_time <- parse_date_time(string_of_time, c("%Y.%m.%d %H:%M", "%m.%d.%Y %H:%M", "%Y.%m.%d %H:%M:%S"))
str(string_of_time_as_time)

```

[@lubridate].

### Parts of time

For *ms*, the package *lubridate* is needed. This function extracts the minutes and seconds.

```{r label = "ms"}
time <- "2:53"
parsed_time <- ms(time)
print(parsed_time)

```

The time in the last minute is $\qty{`r parsed_time`}{\s}$.

```{r label = "ms-as-numeric"}
s <- as.numeric(parsed_time)

```

The time is $\qty{`r s`}{\s}$.

## Array

### Creation

An array can be created using the function `c`:

```{r label = "array"}
array <- c("value_1", "value_2")
print(array)

```

### Referencing position

Referencing a position inside the array takes place using brackets whereas the first position has the index `1`:

```{r label = "array-referencing"}
value_at_position_2 <- array[2]
print(value_at_position_2)

array[2] <- "value_3"
print(array)

```

### Multidimensional

Arrays with the same length can be connected into `data.frame`:

```{r label = "data.frame"}
array_2 <- c("value_3", "value_4")
data_frame <- data.frame(array, array_2)
print(data_frame)

```

### Column names

A `data.frame` can have column names:

```{r label = "colnames"}
colnames(data_frame) <- c(
  "array",
  "array 2"
)

print(data_frame)

```

### Adding column

```{r label = "cbind"}
cbind(data_frame, array_2)

```

### Row names

```{r label = "rownames"}
rownames(data_frame) <- c(
  "Caption of first row",
  "Caption of second row"
)

print(data_frame)

```

### Importing data

Data can be imported from a text file into a data frame:

```{r label = "read-table", eval = FALSE}
CO2_in_air <- read.table("co2_brw_surface-insitu_1_ccgg_DailyData.txt", header = TRUE, sep = "", dec = ".")

```

Data can also be imported from a comma-separated-values-(CSV-)file into a data frame:

```{r label = "import-washing-cycles"}
washing_cycles <- read.table("washing-cycles.csv", header = TRUE, sep = ",", dec = ".")

```

### Printing first rows

```{r label = "head"}
print(head(washing_cycles))

```

### Editing the look of a cell

```{r label = "cell-spec"}
librarian::shelf("kableExtra")
coloured_cell <- cell_spec(data_frame[, 2], color = "red")

```

[@cell_spec].

### Editing the content

```{r label = "mutate"}
librarian::shelf("dplyr")
mutate(data_frame, "array 2" = "mutated")

```

[@create].

### Size

```{r label = "length"}
length(data_frame)

nrow(data_frame)

```

### Looping

```{r label = "for"}

for (index in 1:length(data_frame)) {
  print(data_frame[index])
}

```

### Referencing by column name

```{r label = "bug"}
print(data_frame$`array 2`)

```

[@r_language].

### Subsetting

As `washing_cycles` also contains records with missing data i want them removed:

```{r label = "subset"}

washing_cycles_with_full_records <- subset(washing_cycles, !is.na(`Veenäit.enne`) & "" != `Veenäit.pärast` & !is.na(`Veekulu..l.`) & !is.na(`kWh`))

print(head(washing_cycles_with_full_records))

```

[@subset].

This was how to remove incomplete records by manually setting the columns that contain empty records. There is a move convenient method to do that without specifying columns:

```{r label = "complete"}
washing_cycles_with_full_records <- washing_cycles[complete.cases(washing_cycles), ]

```

[@r_language].

i only want to see the data in the column `Kava`:

```{r label = "subset-column"}
program_in_washing_cycles <- subset(washing_cycles_with_full_records, select = `Kava`)

print(head(program_in_washing_cycles))

```

i only want to see cycles from the rows 2 to 4 in the second column:

```{r label = "subset-limited-cycles"}
print(washing_cycles_with_full_records[2:4, 2])

```

i only want to see hot cycles:

```{r label = "subset-hot-cycles"}
hot_cycles <- subset(washing_cycles_with_full_records, `Temperatuur` > 40)
print(hot_cycles)

```

i only want to see the indices of the cycles at the temperature of $313.15 \times \mathrm{K}$:

```{r label = "subset-cycles-at-40"}
which(washing_cycles_with_full_records[3] == 40)

```

### Sorting

Displaying the indices of the descending sorted values of a vector:

```{r label = "sorting"}
librarian::shelf("dplyr")
desc(subset(head(washing_cycles_with_full_records), select = `Kava`))

```

[@dplyr, p. 7, 21].

Sorting values in ascending order according to the program:

```{r label = "arrange"}
head(arrange(washing_cycles_with_full_records, `Kava`))

```

### Totals

```{r label = "adorn-totals"}
librarian::shelf("janitor")
adorn_totals(dat = head(subset(x = washing_cycles_with_full_records, select = c(`Algus`, `Veekulu..l.`))), where = "row", fill = "", na.rm = TRUE, name = "Kokku", c(`Veekulu..l.`))

```

[@adorn_totals].

## Functions

Functions can be made using the keyword `function`:

```{r label = "function"}
add <- function(first, second, digits = 2) {
  return(signif(first + second, digits = digits))
}

add(first = 123, second = 456) # using the default value 2 for digits

sum <- add(first = 123, second = 456, digits = 1)

```

## Square root

A square root can be calculated using the function `sqrt()`:

```{r label = "sqrt"}
input_for_square_root <- 4
square_root <- sqrt(input_for_square_root)

```

The square root of $\num{`r as.character(input_for_square_root)`}$ is $\num{`r as.character(square_root)`}$.

## Derivation

```{r label = "derivative"}
initial_function <- "x^3 + x^2"
functionToUse <- parse(text = initial_function)

librarian::shelf(c(
  "Ryacas"
))

derivative = D(functionToUse, "x")
string_of_derivative <- deparse(derivative)

```

The derivative of $`r initial_function`$ is $`r string_of_derivative`$.

```{r label = "critical-places"}
equality <- paste(string_of_derivative, "== 0")
print(equality)
print(paste("Solve(", equality, ", x)", sep = ""))
print(y_rmvars(paste("Solve(", equality, ", x)", sep = "")))
critical_places <- yac_str(y_rmvars(paste("Solve(", equality, ", x)", sep = "")))
print(critical_places)
critical_places_as_r <- as_r(critical_places)
print(critical_places_as_r)
critical_solution_1 <- (critical_places_as_r[1])
critical_solution_2 <- critical_places_as_r[2]

```

The critical solutions of $`r initial_function`$ are `r critical_solution_1` and `r critical_solution_2`.

## Comments

Comments can be done with `#`:

```{r label = "comment"}
print(sum) # printing out the value of sum

```

## Strings

Strings can be written using either apostrophes or quotation marks.

For substituting something inside a string, `gsub` can be used [@r]:

```{r label = "gsub"}
gsub("_", "\\_", array, fixed = TRUE)

```

Here, in order to preserve a backslash, it has to be escaped as otherwise, it escapes the underscore. If i'd turn off `fixed`, the function would work like with regular expressions.

## Branching

```{r label = "ifelse"}
ifelse(is.na(NA), "Not Available", "Available")

```

## Table

A table that isn't just in R code but designed and all can be created using `kable` and `kableExtra`. i also build a wrapper function `print_table` for that purpose so that i don't have to rewrite some general things from table to table. i have to pay attention that there can't be any underscores inside the table unless they are part of an eqation. My sample table is table \@ref(tab:table).

```{r label = "table"}
print_table(
  table = sapply(data_frame, function(value) gsub("_", "\\_", value, fixed = TRUE)),
  caption = "Caption."
)

```

If i want to apply a function to all the values inside a data frame, i use `sapply` [@lathiya_2022_sapply]:

```{r label = "sapply"}
sapply(data_frame, function(value) gsub("_", "\\_", value, fixed = TRUE))

```


Tables \@ref(tab:table-with-digits-default) and \@ref(tab:table-with-digits-nondefault) are for comparing the number of digits after comma. The table \@ref(tab:table-with-digits-default) has the default number of digits and the table \@ref(tab:table-with-digits-nondefault) has another number of digits in every number after comma.

```{r label = "table-with-digits-default"}
water_report <- head(subset(x = washing_cycles_with_full_records, select = c(`Algus`, `Veenäit.enne`, `Veenäit.pärast`)))

librarian::shelf(c(
  'dplyr'
))

water_report <- water_report %>%
  mutate(`Veenäit.pärast` = as.numeric(`Veenäit.pärast`))

colnames(water_report) <- c(
  "Start",
  "$\\mathrm{\\frac{\\text{Used water before}}{\\mathrm{m^3}}}$",
  "$\\mathrm{\\frac{\\text{Used water after}}{\\mathrm{m^3}}}$"
)

print_table(
  table = water_report,
  caption = "Water report with numbers with up to two digits after comma."
)

```

[@dplyr, p. 38 to 42].

Inside the table, backslashes must be escaped.

```{r label = "table-with-digits-nondefault"}
print_table(
  table = water_report,
  caption = "Water report with numbers with up to four digits after comma.",
  digits = 4
)

```

It's also possible for the table to have an additional header whose columns span over multiple columns in the first header [@add_header_above]:

```{r label = "table-with-additional-header"}
print_table(
  table = water_report,
  caption = "Water report with additional header.",
  additional_header = c("Spanned header" = 4)
)

```

It's possible to change the look of a row:

```{r label = "table-with-row-specification"}
print_table(
  table = water_report,
  caption = "Water report with coloured row."
) %>%
  row_spec(2, color = "teal")

```

Here, `%>%` means piping.

And it's possible to change the look of a column:

```{r label = "table-with-column-specification"}
print_table(
  table = water_report,
  caption = "Water report with wider column."
) %>%
  column_spec(1, width = "16em")

```

And what if the table is too wide:

```{r label = "table-with-landscape"}
print_table(
  table = water_report,
  caption = "Water report as landscape."
) %>%
  landscape()

```

Linked footnotes don't work with `kable`. Footnotes can be created like this:

```{r label = "footnotes"}

DATA_FRAME_OF_COMPARISON <- data.frame(
  0.3
)

colnames(DATA_FRAME_OF_COMPARISON) <- c(
  paste("$\\frac{T_\\text{dew}}{}$", footnote_marker_number(1))
)

print_table(
  table = DATA_FRAME_OF_COMPARISON,
  caption = "Water report with a footnote.",
  footnotes = c(
    "juhendi tabel 5.1" # 1
  )
)

```

`threeparttable` must be set to `TRUE` for just in case the footnote is too long for the width of the paper [@footnote].

By default, i feed one-dimensional arrays to data frame and the values of these arrays will be displayed from to to down. If i want them to be displayed from left to right, i have to transform the table:

```{r label = "t"}

DATA_FRAME_OF_COMPARISON <- data.frame(
  0.3
)

rownames(DATA_FRAME_OF_COMPARISON) <- c(
  "\"Pasco\" ilmajaam"
)

colnames(DATA_FRAME_OF_COMPARISON) <- c(
  paste("$\\frac{T_\\text{dew}}{}$", footnote_marker_number(1))
)

print_table(
  DATA_FRAME_OF_COMPARISON,
  caption = "Table with rows and columns exchanged.",
  do_i_transpose = TRUE
)

```

## Exponents and logarithms

$$e^1$$

```{r label = "e"}
exp(1)

```

$$\mathrm{ln}(e)$$

```{r label = "log"}
log(exp(1))

```


## Plotting

### One graph

An example of plotting data is putting data points to the plot - visible on the figure \@ref(fig:ggplot) on the page `r render_pageref(reference = "fig:ggplot")`. The packages:

`ggplot2`
: for plotting [@ggplot2]

`latex2exp`
: for using `r render_nonmath_LaTeX_string_universal("\\LaTeX")` strings in labels

The parameters:

`data`
: the data table

`aes`
: the function that describes axis

`x`
: data array for *x* axis

`y`
: data array for *y* axis

`geom_point`
: the function for plotting data points

`shape`
: the shape of points

`size`
: the size of points

`color`
: the border color

`fill`
: the fill color

`labs`
: the function for creating labels for axis

`TeX`
: the function for converting `r render_nonmath_LaTeX_string_universal("\\LaTeX")` strings in labels

```{r label = "ggplot", fig.cap="Washing cycles."}

librarian::shelf(c(
  "ggplot2",
  "latex2exp"
))

ggplot(data = washing_cycles_with_full_records, aes(x = `Temperatuur`, y = `Pöördeid.min`)) + geom_point(shape = 23, color = "#008000", fill = "#ff6600", size = 3) +
  labs(x = TeX("$\\frac{\\textit{t}}{\\degree C}$"), y = TeX("$\\frac{\\textit{f}}{min}$"))

```

### Multiple graphs

Multiple graphs can be displayed on a same figure as seen on the figure \@ref(fig:ggplot-multiple-graphs) on the page `r render_pageref(reference = "fig:ggplot-multiple-graphs")`.

```{r label = "ggplot-multiple-graphs", fig.cap=paste("Water and electricity consumption between", min_time, "and", max_time, ".")}
librarian::shelf(c(
  "lubridate"
))

washing_cycles_with_full_records <- washing_cycles_with_full_records %>%
  mutate(`Algus` = parse_date_time(x = washing_cycles_with_full_records$Algus, orders = c(
  "%Y-%m-%d %H",
  "%Y.%m.%d %H:%M",
  "%Y.%m.%d %H:%M:%S",
  "%m.%d.%Y %H:%M",
  "%m.%d.%Y %H:%M:%S"
))) %>%
   mutate(`veekulu_dal` = (as.numeric(washing_cycles_with_full_records$Veenäit.pärast) - washing_cycles_with_full_records$Veenäit.enne) * 100)

min_time = min(washing_cycles_with_full_records$Algus)
max_time = max(washing_cycles_with_full_records$Algus)

librarian::shelf(c(
  "ggplot2",
  "latex2exp"
))

ggplot(data = washing_cycles_with_full_records, mapping = aes(x = `Algus`, y = `veekulu_dal`)) +
  geom_point(shape = 23, color = "#008000", fill = "#008000", size = 3) +
  labs(x = "Algus", y = TeX("$\\frac{\\textit{V} \\times 10^2}{m^3}$")) +
  geom_point(mapping = aes(x = `Algus`, y = `kWh`), color = "#ff6600", fill = "#ff6600", shape = 24, size = 3)  +
  scale_y_continuous(sec.axis = sec_axis(~., name = TeX("\\frac{\\textit{W}}{kWh}"))) +
  theme(axis.title.y = element_text(colour = "#008000"), axis.text.y = element_text(colour = "#008000"), axis.ticks.y = element_line(colour = "#008000"), axis.title.y.right = element_text(colour = "#ff6600"), axis.ticks.y.right = element_line(colour = "#ff6600"), axis.text.y.right = element_text(colour = "#ff6600"))

```

### Regression and correlation

The figure \@ref(fig:stat-regline-cor) on the page `r render_pageref(reference = "fig:stat-regline-cor")` represents the correlation between the water and electricity consumption between `r min_time` and `r max_time`. The plot displays the correlation coefficient. There are also the data about the regression. Both labels have been positioned vertically.

```{r label = "stat-regline-cor", fig.cap=paste("Correlation between the water and electricity consumption between", min_time, "and", max_time, ".")}

librarian::shelf(c(
  "ggplot2",
  "ggpubr", # for stat_regline_equation
  "latex2exp"
))

if (!decimal_separator_period) { # for stat_regline_equation and stat_cor
  options(OutDec = ".")
}
ggplot(data = washing_cycles_with_full_records, mapping = aes(x = `Veekulu..l.`, y = `kWh`)) +
  geom_point(shape = 23, color = "#ff6600", fill = "#ff6600", size = 3) +
  labs(x = TeX("$\\frac{\\textit{V}}{m^3}$"), y = TeX("$\\frac{\\textit{W}}{kWh}$")) +
  stat_regline_equation(mapping = aes(x = `Veekulu..l.`, y = `kWh`, label = paste(..eq.label.., ..rr.label.., ..adj.rr.label.., sep = "~~~~")), color = "#ff6600", label.y = 3) +
  stat_cor(aes(x = `Veekulu..l.`, y = `kWh`), color = "#ff6600", label.y = 2.5)
if (!decimal_separator_period) {
  options(OutDec = ",")
}

```

### Error bars

The figure \@ref(fig:ggplot-errorbar) on the page `r render_pageref(reference = "fig:ggplot-errorbar")` represents the water and electricity consumption between `r min_time` and `r max_time` with errorbars.

```{r label = "ggplot-errorbar", fig.cap=paste("Water and electricity consumption between", min_time, "and", max_time, " with errobars.")}

librarian::shelf(c(
  "ggplot2",
  "latex2exp"
))

margin_of_V <- 1e-3 / 2 * 2 * 1e2
margin_of_W <- .1 / 2

color_x <- "#008000"
color_y <- "#ff6600"

ggplot(data = washing_cycles_with_full_records, mapping = aes(x = `veekulu_dal`, y = `kWh`)) +
  geom_point(shape = 23, size = 1) +
  labs(x = TeX("$\\frac{\\textit{V} \\times 10^2}{m^3}$"), y = TeX("$\\frac{\\textit{W}}{kWh}$")) +
  geom_errorbarh(aes(xmin = `veekulu_dal` - margin_of_V, xmax = `veekulu_dal` + margin_of_V, y = `kWh`), color = color_x) +
  geom_errorbar(aes(x = `veekulu_dal`, ymin = `kWh` - margin_of_W, ymax = `kWh` + margin_of_W), color = color_y) +
  theme(axis.title.x = element_text(colour = color_x), axis.text.x = element_text(colour = color_x), axis.ticks.x = element_line(colour = color_x), axis.title.y = element_text(colour = color_y), axis.text.y = element_text(colour = color_y), axis.ticks.y = element_line(colour = color_y))

```
